A point-level trend model for the North American Breeding Bird Survey that corrects for detectability

### 2024-10-28 article/data-chapter-3.tex
changes for thesis

Given the recent detectability offsets generated by the NA-POPS project \citep{edwards_point_2023}, it would be useful to be able to incorporate these detectability offsets into the current BBS status and trends model. 
Incorporating these offsets can serve several purposes, from correcting for roadside and landcover biases via statistical offsets \citep{thogmartin_sensitivity_2010, solymos_lessons_2020, edwards_point_2023}, to integrating disparate data sources \citep{solymos_calibrating_2013, edwards_point_2023}.
Additionally, trends will have included the effects of changing detectability through time, because this source of variation will be accounted for through these offsets.
However, because the current BBS model considers summed counts at the route level (rather than at the individual point level), we cannot yet directly apply these offsets to BBS data.

\par In this paper, we describe an extension of the current BBS status and trend model \citep{smith_spatially_2023} that considers BBS data at the point-level, to allow for detectability to be explicitly included in trend analysis. 
Using point-level (also called stop-level) data, we compare trends and trajectories of Ovenbird (\textit{Seiurus aurocapilla}) obtained through this point-level model to trends and trajectories obtained through the current route-level BBS models described in \citet{smith_spatially_2023}.
After establishing this baseline comparison, we incorporate detectability offsets into the new point-level model in two ways: first using a naive approach that considers these detectability offsets as fixed values, and secondly by incorporating error associated with the derived detectability offsets and propagating forward the uncertainty using variance propagation methods described in \citet{bravington_variance_2021}.
Finally, we investigate how incorporating detectability as offsets affects the trends and trajectories for Ovenbird across Canada, and in particular focus on areas where there have been large landcover changes.

@@

\par The following subsections describe how the current route-level description of the BBS status and trend model can be converted to a point-level model, by taking advantage of the point-level data provided by the United States Geological Survey and Canadian Wildlife Service that has been available since 1997. 
We first describe the current status and trend model used by the Canadian Wildlife Service, as well as relevant Bayesian priors and assumptions with that model.
We then describe how to use the point-level data in this status and trend model, as well as describe additional choices a researcher could make regarding route-level hyperparameters.
Then, we expand this point-level model to explicitly include detectability estimates, as well as uncertainty around those estimates using variance propagation techniques described by \citep{bravington_variance_2021}.
Table \ref{params_table} provides a reference guide to all the parameters used in these models.

\begin{table}[!ht]
	\caption{Parameters in the BBS models. Equation \ref{bbs} refers to the route-level model; Equation \ref{bbs-point} refers to the point-level model; Equation \ref{bbs-detect} refers to the point-level model with detectability offsets; and Equation \ref{bbs_varprop} refers to the point-level model with detectability and variance propagation.}
	\label{params_table}
		\begin{tabular}[t]{l>{\raggedright}p{0.5\linewidth}>{\raggedright\arraybackslash}p{0.3\linewidth}}
		\hline
		Parameter or Datum&Description&Model\\
		\hline
		$i$&Stratum index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$j$&Point/stop index&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$J$&Route index&Equation \ref{bbs}\\
		$t$&Year index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$o$&Observer index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$Y_{J,i,t,o}$&Total observed birds at route $J$, stratum $i$, year $t$, by observer $o$&Equation \ref{bbs}\\
		$Y_{j,i,t,o}$&Total observed birds at point $j$, stratum $i$, year $t$, by observer $o$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\alpha_i$&Stratum-level intercept, modelled with an intrinsic conditional autoregressive model \citep{besag_bayesian_1991, smith_spatially_2023}&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\delta_{i,t}$&Smoothed temporal component estimating population trajectory through time. Modelled with generalized additive model \citep{smith_north_2021}&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\gamma_{i,t}$&Annual year-effect fluctuation term; departure from smoothed GAM&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\psi_J$&Route-level intercept&Equation \ref{bbs}\\
		$\psi_j$&Point-level intercept&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\omega_o$&Observer-level intercept&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\eta$&First-year observer effect&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$I(o,J,t)$&1 if observer is surveying route $J$ in year $t$ for the first time, 0 otherwise&Equation \ref{bbs}\\
		$I(o,j,t)$&1 if observer is surveying point $j$ in year $t$ for the first time, 0 otherwise&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$p_{j,i,t}$&Probability of availability at point $j$ in stratum $i$ in year $t$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$q_{j,i,t}$&Probability of perceptibility at point $j$ in stratum $i$ in year $t$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		\hline
	\end{tabular}
\end{table}

\begin{table}[!ht]
	\caption{Table \ref{params_table} continued}
	\begin{tabular}[t]{l>{\raggedright}p{0.5\linewidth}>{\raggedright\arraybackslash}p{0.3\linewidth}}
		\hline
		Parameter or Datum&Description&Model\\
		\hline
		$\kappa_{j,i,t}^{(p)}$&Design matrix for availability variance propagation; first derivatives of probability of availability \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		$\kappa_{j,i,t}^{(q)}$&Design matrix for perceptibility variance propagation; first derivatives of probability of perceptibility \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		$\zeta$&Effects vector for availability variance propagation \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		$\xi$&Effects vector for perceptibility variance propagation \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		\hline
	\end{tabular}
\end{table}

@@

\begin{equation*}
	Y_{J,i,t,o} | \alpha_i, \delta_{i,t}, \gamma_{i,t}, \psi_J, \omega_o, \eta \sim \text{NegBin}(\lambda_{J,i,t,o}, \nu)
\end{equation*}
\begin{equation}\label{bbs}
\log(\lambda_{J,i,t,o}) = \alpha_i + \delta_{i,t} + \gamma_{i,t} + \psi_J + \omega_o + \eta I(o,J,t) + \epsilon_{J,i,t,o}.
\end{equation}
Counts $Y$ for a given species on each route $J$ in stratum $i$, in year $t$ conducted by observer $o$ are modelled as realizations of a negative binomial distribution with mean $\lambda_{J,i,t,o}$ and inverse dispersion parameter $\nu$ \citet{smith_spatially_2023}. 
Note that we have explicitly chosen to use capital letter $J$ to represent the route, as this will be differentiated from specific points $j$ within a route $J$ later in this paper.
On the log scale, values of $\lambda$ are modelled by intercepts representing mean count for each stratum ($\alpha_i$), route($\psi_J$), and observer ($\omega_o$), plus a temporal component $\delta_{i,t}$ that estimates population trajectory through time. 
The stratum intercept $\alpha_i$ are estimated via intrinsic conditional autoregressive (ICAR) models \citep{besag_bayesian_1991}, allowing for information to be shared explicitly in space \citep{smith_spatially_2023}.
Since the temporal component is a smoothed GAM \citep{smith_north_2021}, we add in an annual fluctuation term $\gamma_{i,t}$ to allow for fluctuations from the smooth. 
We also have a first-year observer term $\eta$ that is present if the route $J$ in year $t$ is run by observer $o$ for the first time, and is zero otherwise. 
Finally, we also have a random noise term $\epsilon_{J,i,t,o}$.
Because this model is estimated in a Bayesian context, we set priors on these parameters based on previous work from \citet{smith_north_2021} and \citet{smith_spatially_2023}.

@@

\par Recall that the model described in Equation \ref{bbs} sums all points within a route $J$ and models these summed counts for each route $J$, such that the term $\psi_J$ is the route-level random effect and $\log(\lambda_{j,i,t,o})$ represents the mean expected count at route $J$ in stratum $i$ during year $t$ by observer $o$.
Rather than modelling at the route-level $J$, we can, if we treat each point along a route as a random sample, consider each individual point or ``stop" within a route as a count.
We then be model the data at the point-level $j$.
The term $\psi_j$ then represents a point-level random effect (as opposed to a route-level random effect in Equation \ref{bbs}), and $\log(\lambda_{j,i,t,o})$ would represent the mean expected count at point $j$ (rather than route $J$) in stratum $i$ during year $t$ by observer $o$. 
Our point-level model now becomes

>>>>

Given the detectability offsets generated by the NA-POPS project \citep{edwards_point_2023}, it would be useful to be able to incorporate these detectability offsets into the current BBS status and trends model. 
Incorporating these offsets can serve several purposes, from correcting for roadside and landcover biases via statistical offsets \citep{thogmartin_sensitivity_2010, solymos_lessons_2020, edwards_point_2023}, to integrating disparate data sources \citep{solymos_calibrating_2013, edwards_point_2023}.
Additionally, trends will have included the effects of changing detectability through time, because this source of variation will be accounted for through these offsets.
However, because the current BBS model considers summed counts at the route level (rather than at the individual point level), we cannot yet directly apply these offsets to BBS data.

\par In this chapter, I describe an extension of the current BBS status and trend model \citep{smith_spatially_2023} that considers BBS data at the point-level, to allow for detectability to be explicitly included in trend analysis. 
Using point-level (also called stop-level) data, I compare trends and trajectories of Ovenbird (\textit{Seiurus aurocapilla}) obtained through this point-level model to trends and trajectories obtained through the current route-level BBS models described in \citet{smith_spatially_2023}.
After establishing this baseline comparison, I incorporate detectability offsets into the new point-level model in two ways: first using a naive approach that considers these detectability offsets as fixed values, and secondly by incorporating error associated with the derived detectability offsets and propagating forward the uncertainty using variance propagation methods described in \citet{bravington_variance_2021}.
Finally, I investigate how incorporating detectability as offsets affects the trends and trajectories for Ovenbird across Canada, and in particular focus on areas where there have been large landcover changes.

@@

\par The following subsections describe how the current route-level description of the BBS status and trend model can be converted to a point-level model, by taking advantage of the point-level data provided by the United States Geological Survey and Canadian Wildlife Service that has been available since 1997. 
I first describe the current status and trend model used by the Canadian Wildlife Service, as well as relevant Bayesian priors and assumptions with that model.
I then describe how to use the point-level data in this status and trend model, as well as describe additional choices a researcher could make regarding route-level hyperparameters.
Then, I expand this point-level model to explicitly include detectability estimates, as well as uncertainty around those estimates using variance propagation techniques described by \citep{bravington_variance_2021}.
Table \ref{table:params-bbs} provides a reference guide to all the parameters used in these models.

	\begin{longtable}{>{\hspace{0pt}}m{0.2\linewidth}>{\hspace{0pt}}m{0.5\linewidth}>{\hspace{0pt}}m{0.3\linewidth}}
	\caption{Parameters in the BBS models. Equation \ref{bbs} refers to the route-level model; Equation \ref{bbs-point} refers to the point-level model; Equation \ref{bbs-detect} refers to the point-level model with detectability offsets; and Equation \ref{bbs_varprop} refers to the point-level model with detectability and variance propagation.\label{table:params-bbs}}\\ 
	\hline
	Parameter or Datum~                                          & Description~                                  & Model~                                                                                                                                                                                                                                                                                                                                             \endfirsthead 
	\hline
	$i$&Stratum index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$j$&Point/stop index&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$J$&Route index&Equation \ref{bbs}\\
	$t$&Year index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$o$&Observer index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$Y_{J,i,t,o}$&Total observed birds at route $J$, stratum $i$, year $t$, by observer $o$&Equation \ref{bbs}\\
	$Y_{j,i,t,o}$&Total observed birds at point $j$, stratum $i$, year $t$, by observer $o$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$\alpha_i$&Stratum-level intercept, modelled with an intrinsic conditional autoregressive model \citep{besag_bayesian_1991, smith_spatially_2023}&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$\delta_{i,t}$&Smoothed temporal component estimating population trajectory through time. Modelled with generalized additive model \citep{smith_north_2021}&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$\gamma_{i,t}$&Annual year-effect fluctuation term; departure from smoothed GAM&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$\psi_J$&Route-level intercept&Equation \ref{bbs}\\
	$\psi_j$&Point-level intercept&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$\omega_o$&Observer-level intercept&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$\eta$&First-year observer effect&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$I(o,J,t)$&1 if observer is surveying route $J$ in year $t$ for the first time, 0 otherwise&Equation \ref{bbs}\\
	$I(o,j,t)$&1 if observer is surveying point $j$ in year $t$ for the first time, 0 otherwise&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$p_{j,i,t}$&Probability of availability at point $j$ in stratum $i$ in year $t$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$q_{j,i,t}$&Probability of perceptibility at point $j$ in stratum $i$ in year $t$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
	$\kappa_{j,i,t}^{(p)}$&Design matrix for availability variance propagation; first derivatives of probability of availability \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
	$\kappa_{j,i,t}^{(q)}$&Design matrix for perceptibility variance propagation; first derivatives of probability of perceptibility \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
	$\zeta$&Effects vector for availability variance propagation \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
	$\xi$&Effects vector for perceptibility variance propagation \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
	\hline
\end{longtable}

@@

\begin{equation*}
	Y_{J,i,t,o} | \alpha_i, \delta_{i,t}, \gamma_{i,t}, \psi_J, \omega_o, \eta \sim \text{NegBin}(\lambda_{J,i,t,o}, \nu)
\end{equation*}
\begin{equation}\label{bbs}
\log(\lambda_{J,i,t,o}) = \alpha_i + \delta_{i,t} + \gamma_{i,t} + \psi_J + \omega_o + \eta I(o,J,t) + \epsilon_{J,i,t,o}.
\end{equation}
Counts $Y$ for a given species on each route $J$ in stratum $i$, in year $t$ conducted by observer $o$ are modelled as realizations of a negative binomial distribution with mean $\lambda_{J,i,t,o}$ and inverse dispersion parameter $\nu$ \citet{smith_spatially_2023}. 
Note that I have explicitly chosen to use capital letter $J$ to represent the route, as this will be differentiated from specific points $j$ within a route $J$ later in this paper.
On the log scale, values of $\lambda$ are modelled by intercepts representing mean count for each stratum ($\alpha_i$), route($\psi_J$), and observer ($\omega_o$), plus a temporal component $\delta_{i,t}$ that estimates population trajectory through time. 
The stratum intercept $\alpha_i$ are estimated via intrinsic conditional autoregressive (ICAR) models \citep{besag_bayesian_1991}, allowing for information to be shared explicitly in space \citep{smith_spatially_2023}.
Since the temporal component is a smoothed GAM \citep{smith_north_2021}, we add in an annual fluctuation term $\gamma_{i,t}$ to allow for fluctuations from the smooth. 
We also have a first-year observer term $\eta$ that is present if the route $J$ in year $t$ is run by observer $o$ for the first time, and is zero otherwise. 
Finally, we also have a random noise term $\epsilon_{J,i,t,o}$.
Because this model is estimated in a Bayesian context, we set priors on these parameters based on previous work from \citet{smith_north_2021} and \citet{smith_spatially_2023}.

@@

\par Recall that the model described in Equation \ref{bbs} sums all points within a route $J$ and models these summed counts for each route $J$, such that the term $\psi_J$ is the route-level random effect and $\log(\lambda_{j,i,t,o})$ represents the mean expected count at route $J$ in stratum $i$ during year $t$ by observer $o$.
Rather than modelling at the route-level $J$, we can, if we treat each point along a route as a random sample, consider each individual point or ``stop" within a route as a count.
We would then be model the data at the point-level $j$.
The term $\psi_j$ then represents a point-level random effect (as opposed to a route-level random effect in Equation \ref{bbs}), and $\log(\lambda_{j,i,t,o})$ would represent the mean expected count at point $j$ (rather than route $J$) in stratum $i$ during year $t$ by observer $o$. 
Our point-level model now becomes
